Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-05 paper/paper.tex
implemented kofke's changes and removed quaternions

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{../figures/}}

@@

In this article, a configuration refers to the particles, coordinates and periodic boundaries.
For an $n$-component system, the total number of particles, $N=\sum_{i}^{n} N_i$, where $N_i$ is the number of particles of type $i$.
In bulk systems, there are no interactions of the particles with boundaries.
To avoid spurious effects of walls or a container, and thereby minimizing the size of the simulated system needed to model the bulk, periodic boundary conditions are imposed such that particles interact with their nearest neighboring image \cite{allen_computer_1989}.
To simplify the notation in the following derivations, the periodic boundaries align with the Cartesian axes to form a cuboid of $D$ linear dimensions $L_d$, where the volume, $V$, is given by $V=\prod_d^D L_d$.
The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the interactions could be orientationally-dependent.
But if the particle is not isotropic or rigid, the derivations in this article leave a placeholder variable for the orientation contributions that are not further detailed and are beyond the scope of this article.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\label{sec:lhs_rotation}Particle orientation change trial}
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Orientations are typically fixed during particle displacements so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the displacement parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move which proceeds as follows.
A particle is randomly chosen among $N_m$ mobile particles.
A new orientation of that particle is then chosen according to the probability $P_{\omega n}\diff\boldsymbol{\omega}$.
The transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
 \hline
 \thead{Forward} & \thead{$\alpha_{o\rightarrow n}$} \\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose orientation} & \makecell{$P_{\omega n}\diff\boldsymbol{\omega}$} \\
 \hline\hline
 \thead{Reverse} & \thead{$\alpha_{n\rightarrow o}$}\\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose orientation} & \makecell{$P_{\omega o}\diff\boldsymbol{\omega}$} \\
 \hline
\end{tabular}
\caption{Orientation-change transition probabilities.}
\label{tab:lhs_rotation}
\end{center}
\end{table}

Substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance} and using the transition probabilities in Table~\ref{tab:lhs_rotation} yields,
\begin{equation}
\chi = \frac{P_{\omega o}}{P_{\omega n}}e^{-\beta\Delta U}
\label{eq:lhs_rotate}
\end{equation}
for the Metropolis acceptance of particle orientation changes.
For a rigid rotation of a particle in three dimensions, the orientation may be given by Euler angles or quaternions.
However, when obtaining a random orientation, it is important not to select random values uniformly within the range of each of these variables.
For example, independent and uniform selection of Euler angles will bias orientations toward the poles and makes the calculation of $P_{\omega n}$ and $P_{\omega o}$ nontrivial.
Instead, the orientation should be uniformly chosen on the surface of a sphere \cite{allen_computer_1989}.
In that case, $P_{\omega n}=P_{\omega o}$.
The d$\boldsymbol{\omega}$ terms in the probability of states, $\Pi$, include the orientations of the particles, which are sampled uniformly over the entire surface of a sphere and are normalized with the appropriate Jacobian.
An example of randomly picking points on the surface of a sphere is shown in Section~\ref{sec:unit_sphere}.

A more efficient approach for sampling particle orientations is to perturb the rigid orientation using a tunable parameter, $\delta$.
This must also be done in a careful way and not simply by changing the Euler angles uniformly and independently.
One example algorithm to perturb orientation proceeds as follows.
Generate a random quaternion, $\mathbf{q}$, uniformly on the surface of a sphere \cite{vesely_angular_1982}.
If the original particle orientation is represented by a quaternion, $\mathbf{q}_o$, then the newly proposed orientation is given by the quaternion $\mathbf{q}_n=\frac{\mathbf{q}_o+\delta\mathbf{q}}{|\mathbf{q}_o+\delta\mathbf{q}|}$.
If the original particle orientation is represented by a collection of coordinates, use $\mathbf{q}_o=[0, 0, 0, 1]$ (which corresponds to an identity rotation matrix) to generate a rotation matrix from $\mathbf{q}_n$ to perturb the coordinates.
Because of the use of the tunable parameter, $P_{\omega n}$ and $P_{\omega o}$ are equal to each other and therefore cancel one another in Eq.~\ref{eq:lhs_rotate}.

A larger tunable parameter, $\delta$, leads to a larger rotation, and thus $\delta$ may be tuned in the same manner as the displacement parameter to reach a target acceptance for each species.
To avoid violating detailed balance, this tunable parameter should remain constant during the production simulations for the same reason as shown in Fig.~\ref{fig:lhs_nvt}.
An example of randomly picking quaternions, and thus orientations, is shown in Section~\ref{sec:unit_sphere}.
For particles with multiple interaction sites, one site is typically chosen as the center of the rigid body rotation.
Because picking this one fixed site is deterministic and does not involve a random number generator, there is no associated probability that factors into Eq.~\ref{eq:lhs_rotate}.

For particles with multiple sites that are flexibly bonded, if the particle is not rigidly rotated but instead attempted to be regrown with configurational bias, starting from a given position of one of the interaction sites, then $P_{\omega n}$ and $P_{\omega o}$ are related to the product of the probability of choosing each of those interaction site positions \cite{rosenbluth_monte_1955, frenkel_understanding_2002}.
Particle regrowth is the subject of many studies \cite{martin_novel_1999, martin_using_2006, sepehri_improving_2017, sepehri_improving_2017-1}, but is beyond the scope of this article, which concerns only the placement of the center of the first bead.

@@

Although the derivation of the acceptance criteria for MC trials is the focus of this article, common errors encountered during the implementation of MC are included here.
For a more comprehensive discussion of the inner workings of MC see Ref.~\cite{dubbeldam_inner_2013}.
Thus, the scope of this section is for testing MC trial implementations.
In this section, we discuss energy tests and machine precision issues for helping to capture common errors, the calculation of standard deviations using the block method \cite{flyvbjerg_error_1989, grossfield_best_2018}, the generation of a random position on the surface of a unit sphere and spherical shell, and the generation of random quaternions and rotation matrices.
This section then ends with a discussion of the general workflow for testing a new MC trial.

@@

In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
However, one must be careful in how an orientation is generated.
Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random point on the surface of a sphere \cite{vesely_angular_1982}.
Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
A randomly generated perturbation of orientation is then demonstrated in Fig.~\ref{fig:random_orientation}.

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:quaternion},
caption=The quaternion module generates a random quaternion about the surface of a sphere in three dimensions.
]{../codes/quaternion.py}
\end{figure}

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{soul}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{../figures/}}

@@

In this article, a configuration refers to the particles, coordinates and periodic boundaries.
For an $n$-component system, the total number of particles, $N=\sum_{i}^{n} N_i$, where $N_i$ is the number of particles of type $i$.
In bulk systems, there are no interactions of the particles with boundaries.
To avoid spurious effects of walls or a container, and thereby minimizing the size of the simulated system needed to model the bulk, periodic boundary conditions are imposed such that particles interact with their nearest neighboring image \cite{allen_computer_1989}.
To simplify the notation in the following derivations, the periodic boundaries align with the Cartesian axes to form a cuboid of $D$ linear dimensions $L_d$, where the volume, $V$, is given by $V=\prod_d^D L_d$.

The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and/or orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the site-site interactions could be orientationally-dependent (e.g., point dipoles).
The orientation of a particle might be given implicitly (not part of $\boldsymbol{\omega}$) via the Cartesian coordinates of its interaction sites.
Alternatively, it may be specified explicitly via an appropriate orientation vector, with the location of the interaction sites (if a multi-site particle) constructed on-the-fly from the position, orientation, and internal coordinates.
In a 3-dimensional space, the orientation coordinate for a cylindrically-symmetric object requires just two independent values, e.g., a unit  vector describing the cylindrical axis direction.
For particles with no such symmetry, three independent values are needed to specify the orientation.
Euler angles ($\phi\theta\psi$) are sometimes invoked for this, but they are not particularly useful in the context of molecular simulation.
A better choice is a quaternion ${\bf q}$, which is a normalized, four-element vector.\cite{?}

If the particle is not isotropic or rigid, the derivations in this article leave a placeholder variable for the orientation contributions that are not further detailed and are beyond the scope of this article.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\label{sec:lhs_rotation}Particle orientation-change trial}
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Orientations are typically fixed during particle displacements so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the displacement parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move.
A robust and intuitive way to accomplish this is via the axis-angle method: (1) a particle is randomly chosen among $N_m$ mobile particles; (2) a rotation axis $\hat {\bf u}$ is generated by sampling uniformly on the surface of a sphere (Sec.~\ref{sec:unit_sphere}); (3) a rotation angle $\theta$ is generated uniformly on $\pm\delta/2$ (with $\delta$ a tunable parameter).
With the rotation so defined, a rotation matrix can be computed
\begin{equation}
\mathbf{R} =
\begin{bmatrix}
C + (1 - C) u_x^2 & (1 - C) u_x u_y - S u_z & (1 - C) u_x u_z + S u_y \\
(1 - C) u_y u_x + S u_z & C + (1 - C) u_y^2 & (1 - C) u_y u_z - S u_x \\
(1 - C) u_z u_x - S u_y & (1 - C) u_z u_y + S u_x & C + (1 - C) u_z^2
\end{bmatrix}
\end{equation}
where $C =\cos\theta$ and $S = \sin\theta$.
The new coordinate of interaction site $j$ on particle $i$ at ${\bf r}_i$ is given by ${{\bf r}_{ij,{\rm n}}={\bf r}_{i}}+{\mathbf R}({\bf r}_{ij,{\rm o}}-{\bf r}_i)$.
If ${\bf r}_i$ is chosen as the centroid of the particle, then the centroid will be unchanged by the rotation (complementing the translation move, which leaves the orientation unchanged).
However, using the centroid for ${\bf r}_i$ is not a requirement; any well-defined center for the rotation may be employed (e.g., the position of a specific interaction site).

%Alternatively, if the orientation coordinate is held explicitly (e.g., as a quaternion), then it can be updated from $\hat{\bf u}$ and $\theta$ using an appropriate formula.

Regardless of how the selected rotation is implemented, the transition probabilities are given in terms of the axis-angle selection process outlined above.
The elements of the transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.
Together they yield the forward and reverse probabilities, which in this case are again equal
\begin{equation}
\alpha_{o\rightarrow n}=\alpha_{n\rightarrow o} = \frac{\diff\mathbf{\omega}\diff\theta}{N_m 4\pi\delta}.
\label{eq:lhs_rhs_rot_fwdrev}
\end{equation}

%Instead, the orientation of a single particle may be perturbed in a separate trial move which proceeds as follows.
%A particle is randomly chosen among $N_m$ mobile particles.
%A new orientation of that particle is then chosen according to the probability $P_{\omega n}\diff\boldsymbol{\omega}$.
%The transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
 \hline
 \thead{Forward} & \thead{$\alpha_{o\rightarrow n}$} \\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
 \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta$} \\
 \hline\hline
 \thead{Reverse} & \thead{$\alpha_{n\rightarrow o}$}\\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
 \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta$} \\
 \hline
\end{tabular}
\caption{Orientation-change transition probabilities.}
\label{tab:lhs_rotation}
\end{center}
\end{table}

Substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance} and using the transition probabilities in Table~\ref{tab:lhs_rotation} yields,
\begin{equation}
\chi = e^{-\beta\Delta U}
\label{eq:lhs_rotate}
\end{equation}
for the Metropolis acceptance of particle orientation changes.
%For a rigid rotation of a particle in three dimensions, the orientation may be given by Euler angles or quaternions.
%However, when obtaining a random orientation, it is important not to select random values uniformly within the range of each of these variables.
%For example, independent and uniform selection of Euler angles will bias orientations toward the poles and makes the calculation of $P_{\omega n}$ and $P_{\omega o}$ nontrivial.
%Instead, the orientation should be uniformly chosen on the surface of a sphere \cite{allen_computer_1989}.
%In that case, $P_{\omega n}=P_{\omega o}$.
%The d$\boldsymbol{\omega}$ terms in the probability of states, $\Pi$, include the orientations of the particles, which are sampled uniformly over the entire surface of a sphere and are normalized with the appropriate Jacobian.
%An example of randomly picking points on the surface of a sphere is shown in Section~\ref{sec:unit_sphere}.
%
%A more efficient approach for sampling particle orientations is to perturb the rigid orientation using a tunable parameter, $\delta$.
%This must also be done in a careful way and not simply by changing the Euler angles uniformly and independently.
%One example algorithm to perturb orientation proceeds as follows.
%Generate a random quaternion, $\mathbf{q}$, uniformly on the surface of a sphere \cite{vesely_angular_1982}.
%If the original particle orientation is represented by a quaternion, $\mathbf{q}_o$, then the newly proposed orientation is given by the quaternion $\mathbf{q}_n=\frac{\mathbf{q}_o+\delta\mathbf{q}}{|\mathbf{q}_o+\delta\mathbf{q}|}$.
%If the original particle orientation is represented by a collection of coordinates, use $\mathbf{q}_o=[0, 0, 0, 1]$ (which corresponds to an identity rotation matrix) to generate a rotation matrix from $\mathbf{q}_n$ to perturb the coordinates.
%Because of the use of the tunable parameter, $P_{\omega n}$ and $P_{\omega o}$ are equal to each other and therefore cancel one another in Eq.~\ref{eq:lhs_rotate}.

A larger tunable parameter, $\delta$, leads to a larger rotation, and thus $\delta$ may be tuned in the same manner as the displacement parameter to reach a target acceptance for each species.
To avoid violating detailed balance, this tunable parameter should remain constant during the production simulations for the same reason as shown in Fig.~\ref{fig:lhs_nvt}.

%If desired, generation of orientations from scratch, without perturbing on an existing orientation, is accomplished by random generation of a quaternion \cite{vesely_angular_1982}.
%An example of randomly picking quaternions, and thus orientations, is shown in Section~\ref{sec:unit_sphere}.

@@

Although the derivation of the acceptance criteria for MC trials is the focus of this article, common errors encountered during the implementation of MC are included here.
For a more comprehensive discussion of the inner workings of MC see Ref.~\cite{dubbeldam_inner_2013}.
Thus, the scope of this section is for testing MC trial implementations.
In this section, we discuss energy tests and machine precision issues for helping to capture common errors, the calculation of standard deviations using the block method \cite{flyvbjerg_error_1989, grossfield_best_2018}, the generation of a random position on the surface of a unit sphere and spherical shell, and the generation of random rotation matrices.
This section then ends with a discussion of the general workflow for testing a new MC trial.

@@

%In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
%This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
%However, one must be careful in how an orientation is generated.
%Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
%Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random point on the surface of a sphere \cite{vesely_angular_1982}.
%Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
%A randomly generated perturbation of orientation is then demonstrated in Fig.~\ref{fig:random_orientation}.
Situations arise where one needs to generate a point at random uniformly on the surface of a sphere in 3D.  This is equivalent to defining an isotropically random direction or a randomly oriented unit vector. There are several ways to approach this: (1) sample the spherical coordinates $\theta$ and $\phi$, appropriately weighted; (2) use a rejection method to generate a point uniformly inside a sphere, i.e., repeatedly sample in a cube that circumscribes the sphere until the sample is also in the sphere, and then normalize the sampled point to a unit vector; (3) sample the $x$, $y$ and $z$ coordinates on independent Gaussian distributions, which yields a point samplied from a spherically symmetric 3D probability density, and normalize to a unit vector. The first method listed here is illustrated in Code Block~\ref{sim:unit_sphere}.

%A similar problem is sampling a quaternion, which may be interpreted as a point on the surface of a unit 4D-hypersphere. Accordingly, Code Block~\ref{sim:quaternion} shows how quaternions may be randomly generated by sampling uniformly on the surface of a 4D hypersphere using an efficient rejection method \cite{vesely_angular_1982}.
%Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices, which could be used to generate a fully random orientation by rotating from an arbitrary reference orientation.
Finally, a randomly generated perturbation of orientation is shown using the axis-angle method in Fig.~\ref{fig:random_orientation}.

@@

%\begin{figure}
%\lstinputlisting[
%language=Python,
%label={sim:quaternion},
%caption=The quaternion module generates a random quaternion about the surface of a 4-dimensional hypersphere.
%]{../codes/quaternion.py}
%\end{figure}
